Refinements for open automata

### 2022-06-30 2021-Refinement/paper/refinementarticle.tex
ajout figure + Defintion 4

\keywords{Labelled transition systems  \and Refinement \and Composition.}
\end{abstract}
%
%
%
\section{Introduction}

Automata are convenient for specifying and verifying systems, but most automata definitions allow only the specifications of finite closed systems. Indeed those systems are the ones we can veify efficiently, but programming often consist in writing open systems that should be interfaced with others, and with potentially unbound behaviours (at least statically).We investigate in our works the possibility to define open symbolic systems, verify some of their properties, and use them to compose in a structured way more complex systems.

Refinement relations verify that and implementation satisfies some properties with respect to a specification. Refinement entails that one system can be considered as a more precise version of the specification, featuring all the specified behaviours with more concrete details. In this article we mostly rely on a simulation point of view, where all the specified behaviour must be followed by the refined system but additionla behaviours may exist. However we also ensure that a while scenario, made of several steps, of the specification can also be simulated by the refined system, which is a bit richer than the traditional simulation relation.

Open automata were defined as a way to provide a semantics for open parameterized hierarchical LTSs used in verification tools.
An open automaton \cite{henrio:01299562} is a classical automaton with variables and holes. Variables makes autoomata symbolic and allow them to encode infinite state systems. Holes enables the composition of automata: an automaton with a hole is an operator that takes another automaton as parameter and reacts to the actions it emits; the composed automaton is a more precise automaton where the behaviour of one ``parameter'' of the main automaton has been provided.

In previous works \cite{2007.10770,wang:03126313} a bisimulation relation was defined for open automata and open parameterized hierarchical LTSs. It has already good properties relatively to bisimulation, but refinement relations were not studied.

This article first introduces open automata that were defined in  \cite{henrio:01299562},  but compared to previous works, we additionally define the composition of open automata.
Then we define  a refinement relation for open automata that has the following characteristics:
\begin{itemize}
\item Classical simulation characterisation but also an additional criteria ensuring that refinement does not introduce deadlocks when following a trace from the simulated automaton.
\item Good properties relatively to composition: we prove here that filling the same hole  with the same automaton preserves the refinement relation.
\item Ability to take into account both composition and transitivity: this is a challenge because composition changes the set of holes of the open automaton and refinement takes into account actions of the holes.
\end{itemize}
The refinement relation is introduced in two steps. First we define a refinement that relates two automata with the same holes, which allows us to focus on the automaton aspect. Second we introduce a relation that relates two automata with different sets of holes, which allows us to take into account the open nature of open automata, and to deal with composition.

@@

@@

\TODO{peut etre echanger c et p? ou c devient p et p devient? }
\begin{definition}[Composition of open automata] \label{Def:CompOA}
The composition of \(A_c = \OAg[c]\) in the hole \(k \in J_p\) of \(A_p = \OAg[p]\) is the OA defined as follows:
\[A_p\subst{A_c}{k} ::=  \OA{S_p \times S_c}{\mpar{s_{0p}, s_{0c}}}{V_p \uplus V_c}{\sigma_{0p} \uplus \sigma_{0c}}{J_c \uplus J_p \setminus \mbrc{k}}{T} \] \text{with }

\begin{align*}
T=& \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s'_c}}{\alpha_p}{\beta_j^{j \in J'_c \uplus J'_p}}{g_p \wedge g_c \wedge \alpha_c = \beta_k}{\psi_p \uplus \psi_c}}{ \OT{s_p}{s'_p}{\alpha_p}{\beta_j^{j \in J'_p \uplus \mbrc{k} }}{g_p}{\psi_p} \in T_p, \OTx{c}{}{}{c} \in T_c} \\
	& \cup \mset{\OT{\mpar{s_p, s_c}}{\mpar{s'_p, s_c}}{\alpha_p}{\beta_j^{j \in J'_p}}{g_p}{\psi_p}}{\OTx{p}{}{}{p} \in T_p, k \notin J'_p, s_c \in S_c}
\end{align*}
\end{definition}
The first OA decides when the second can evolve by involving its hole in a transition:
The action emitted when \(A_c\) makes a transition is synchronised with the action of the hole \(k\) in transitions of \(A_p\) .

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between open automata requires to compare their states. To do so we will suppose that the variables of the two OAs are disjoint (a renaming of variables may have to be applied before comparing OA states)
\begin{definition}[Relation on OA states] Suppose $V_1$ and $V_2$ are disjoint.
A relation on states of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
The idea is that two states are related if the variables they refer to verify a certain formula. Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{022} \vdash R\mpar{s_{01}, s_{02}}\).

@@

>>>>

\keywords{Labelled transition systems  \and Refinement \and Composition.}
\end{abstract}
%
%
%
\section{Introduction}
\TODO{1.5 pages}

Contributions:

We first: introduce open automata (NOT NEW), their composition, 
and the deadlock-free composition.

Then we define a refinement relation for open automata that has the following characteristics:

- good behaviour wrt composition

- refinement does not introduce deadlock

- a first relation that focuses on the automaton part 
and a second that deals with open behaviour

we will see that having at the same time composition and transitivity also raises challenges; we introduce a new form of refinement relation that addresses this challenge.

@@

In Fig. \ref{Fig:SpecOA} we present a very simple example of an open automaton inspired from \cite{Gorrieri:2001}. An automaton that we name \texttt{database} modelling a data base that can be queried and updated. The open automaton \texttt{counter}    depicting  the behaviour of the timer is illustrated in Fig. \ref{Fig:counter}.

\begin{figure}[h]
 \centering
   \includegraphics[width=.9\textwidth]{Figures/database.pdf}
   \caption{Open Automaton. It exposes an unspecified counter in the hole
 to count external tick actions. The system sets a counter when a query is enabled and updates  when  the time is over.\label{Fig:SpecOA}}
\end{figure}

@@

\begin{definition}[Composition of open automata] \label{Def:CompOA}
The composition of \(A_p = \OAg[p]\) in the hole \(k \in J_c\) of \(A_c = \OAg[c]\) is the OA defined as follows:
\[A_c\subst{A_p}{k} ::=  \OA{S_c \times S_p}{\mpar{s_{0c}, s_{0p}}}{V_c \uplus V_p}{\sigma_{0c} \uplus \sigma_{0p}}{J_p \uplus J_c \setminus \mbrc{k}}{T} \] \text{with }

\begin{align*}
T=& \mset{\OT{\mpar{s_c, s_p}}{\mpar{s'_c, s'_p}}{\alpha_c}{\beta_j^{j \in J'_p \uplus J'_c}}{g_c \wedge g_p \wedge \alpha_p = \beta_k}{\psi_c \uplus \psi_p}}{ \OT{s_c}{s'_c}{\alpha_c}{\beta_j^{j \in J'_c \uplus \mbrc{k} }}{g_c}{\psi_c} \in T_c, \OTx{p}{}{}{p} \in T_p} \\
	& \cup \mset{\OT{\mpar{s_c, s_p}}{\mpar{s'_c, s_p}}{\alpha_c}{\beta_j^{j \in J'_c}}{g_c}{\psi_c}}{\OTx{c}{}{}{c} \in T_c, k \notin J'_c, s_p \in S_p}
\end{align*}
\end{definition}

The first OA decides when the second can evolve by involving its hole in a transition:
The action emitted when \(A_p\) makes a transition is synchronised with the action of the hole \(k\) in transitions of \(A_c\).

\begin{figure}[h]
 \centering
   \includegraphics[width=.5\textwidth]{Figures/counter.pdf}\hfill 
   \includegraphics[width=.3\textwidth]{Figures/compose_count_db.pdf}
   \caption{Open Automaton of \texttt{counter} process (left).  Its role is to get a target, then count ticks until that target is reached, then emit an action with the elapsed time and restart. And Open Automation of  the composition \texttt{database}[\texttt{counter}] (right).\label{Fig:counter}}
\end{figure}

\paragraph{Relations between open automata}
A relation (bisimulation, refinement, etc.) between open automata requires to compare their states. To do so we will suppose that the variables of the two OAs are disjoint (a renaming of variables may have to be applied before comparing OA states).
\begin{definition}[Relation on OA states] Suppose $V_1$ and $V_2$ are disjoint.
A relation on states of \(\OAg[1]\) and \(\OAg[2]\) is a function \(R: S_1 \times S_2 \to \rformulas[V_1 \uplus V_2]\).
\end{definition}
The idea is that two states are related if the variables they refer to verify a certain formula. Additionally, we may check that initial states of the automata are related by checking that: \(\sigma_{01} \uplus \sigma_{022} \vdash R\mpar{s_{01}, s_{02}}\).

@@

To illustrate the refinement simulation of open automata, consider a variation on the \texttt{database} automaton of Fig. \ref{Fig:RefineOA}. We suppose that on more concrete level, in addition to querying and updating the data base,  a backup copy of it may also be made.

\begin{figure}[h]
 \centering
   \includegraphics[width=.9\textwidth]{Figures/databaseRefine.pdf}
   \caption{The refined Open Automaton of the automaton of Fig.\ref{Fig:SpecOA}.  Two other actions,  in addition to  \texttt{qry} and \texttt{upd},  \texttt{back} representing a request to make a backup copy and    \texttt{cpy} is used to reflect the actual operation of copying. \label{Fig:RefineOA}}
\end{figure}
